---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SARforecastDLM

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

<!-- lastcommit: start -->
[![Last Commit](https://img.shields.io/github/last-commit/Columbia-Basin-Research-CBR/SARforecastDLM)](https://github.com/Columbia-Basin-Research-CBR/SARforecastDLM/commits/main)
<!-- lastcommit: end -->


**SARforecastDLM** uses Dynamic Linear Modelling (DLM) to forecast changes in one-year ocean survival of wild spring/summer Snake River Chinook salmon, *Oncorhynchus tshawytscha*, from [upwelling indices (CUI)](https://oceanview.pfeg.noaa.gov/products/upwelling/intro), West Coast, USA. This ShinyApp is developed based on the [Scheuerell and Williams (2005)](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1365-2419.2005.00346.x).


## Installation

You can install the development version of **SARforecastDLM** from [GitHub](https://github.com/) with:

``` r
# Install the developmental version of the HydroSurvSizePred package from GitHub
devtools::install_github("Columbia-Basin-Research-CBR/SARforecastDLM")

# Load the package
library(SARforecastDLM)

# Run the app
run_app()
```
**SARforecastDLM** is currently in development and changes are continuously being made. If you have already imported to R studio, please rerun `install_github("Columbia-Basin-Research-CBR/SARforecastDLM")` to see latest changes in the developmental version. If no changes have been made since last import, a warning will appear: `Skipping install of 'SARforecastDLM' from a github remote, the SHA1 (fdd71350) has not changed since last install. Use 'force = TRUE' to force installation`, indicating you have the latest developmental version imported.



If you are interested in the files that support the development version, please see: [https://github.com/Columbia-Basin-Research-CBR/SARforecastDLM](https://github.com/Columbia-Basin-Research-CBR/SARforecastDLM) for files necessary to run. 

The app structure follows a Golem framework described in [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/setting-up-for-success.html) by Colin Fay, Sébastien Rochette, Vincent Guyader and Cervan Girard.

## Contact

This app is being developed by Caitlin O'Brien, Research Scientist, Columbia Basin Research, SAFS, University of Washington. Please reach out with questions/concerns via csobrien@uw.edu.

```{r update_readme, eval=FALSE, include=FALSE}
#You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.
devtools::build_readme()
```

