---
title: "AnotherMarkdownToCleanUp"
output: html_document
date: "2024-05-08"
---

Objective: Add back in removed code with different SAR methods, indices, and potentially passage type and rear type.

Current plan: 
SAR to include: Scheuerell and Williams (2005), CBR DART SAR, CJS estimated survival (Gosselin et al 2019?), and ?
Indices to include: CUI (Bakun index), CUTI, BEUTI?, Northern Copepod Biomass Index
Months to include: April
Passage type to include: All- combined
Rear type to include: Natural-origin


```{r load_libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(MARSS)
library(here)
library(rerddap) #import data from ERDDAP
```
# Northern copepod biomass data: https://www.fisheries.noaa.gov/data-tools/newport-line-northern-copepod-biomass-data ; update with [NOAA](https://www.fisheries.noaa.gov/west-coast/science-data/local-biological-indicators#copepod-biodiversity) referenced data via [Hooff and Peterson, 2006](https://www.fisheries.noaa.gov/west-coast/science-data/local-biological-indicators#copepod-biodiversity); or ask jenn for where to source data
```{r load_cope_data}

#Northern copepod biomass data
cope<-read_csv(here("data-raw", "northern_copepod_biomass_newportOR.csv")) 

cope<-cope %>% 
  filter(Month == 4) %>% 
  group_by(Year) %>% 
  summarise(value = mean(NORTHERN)) %>% 
  select(Year, value) %>% 
  mutate(index = "NCBI")

#load existing sar_raw_data
load(here("data", "sar_raw_data.rda")) #load existing data

## load the salmon sar from MARSS
data(SalmonSurvCUI, package = "MARSS")

#extract sar and year
df<-SalmonSurvCUI %>% 
  select(year, logit.s)

#join sar and year with copepod data
df.cope<-df %>% inner_join( cope, by = c("year" = "Year"))
  
#join sar/cope data with existing sar_raw_data
sar_raw_data_updated<- sar_raw_data %>% 
  bind_rows(df.cope) %>% 
  mutate(sar.method = "Scheuerell and Williams (2005)")

```
# Pull in additional years of CUI
```{r}
### ERDAPP, NOAA upwelling indice CUI
# Set the URL of the ERDDAP server
url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdUI45mo", url = url)

# Download the dataset
dat <- griddap(info,
               time = c("1964-01-01T00:00:00Z", "2024-02-15T00:00:00Z"), #currently pulling all data to date (adjust as needed)
               latitude = c(45,45)
)

# format time and lat/long
df.CUI<-dat$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day), 
         adjlong = longitude-360) #set to -180 to 180

## only include april CUI from ERDAPP data (for now)
df.CUI<-df.CUI %>% 
  filter(month == 4) %>% 
  select(year, upwelling_index) 
```


# pull in additional years of CUTI
```{r}

#load CUTI
# Set the URL of the ERDDAP server
#url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdCUTImonthly", url = url)


# Download the dataset
dat <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2024-02-15T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)


df.CUTI<-dat$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day)) %>% 
  filter(month == 4) %>% 
  select(year, CUTI)  %>% 
  rename(value = "CUTI") %>% 
  mutate(index = "CUTI")



df.cuti<-df %>% inner_join( df.CUTI, by = "year")
  
#join data - returns year and sar repeated for cui and cuti
sar_raw_data<- SalmonSurvCUI %>% 
 rename("value" = CUI.apr) %>% 
  mutate(index = "CUI") %>% 
  bind_rows(df.cuti)

```


# CBR SAR data
```{r}
df.cbr<-read_csv(here("data-raw/SAR", "LGRtoLGA_allpass_W_spsu_ch.csv"))
df.cbr<-df.cbr[-c(24:31),] #drop CBR notes 
df.cbr <-  df.cbr %>% mutate(year = as.integer(year)) %>%   filter( between(year, 2002, 2020)) #2021 does not have ocean4 reported--keep to 2020?


df.cbr<-df.cbr %>% 
  select(year, meanSAR) %>% 
  mutate( logit.s = qlogis(meanSAR/100)) %>% #must divide meanSAR /100 to get 0 to 1 and then apply logit/transform
  select(-meanSAR) 

#CUI
df.cbr.cui<-df.cbr %>% 
  inner_join(df.CUI, by = "year") %>% 
  rename("value" = upwelling_index)  %>%
  mutate(index = "CUI",
         sar.method = "DART")
#CUTI
df.cbr.cuti<-df.cbr %>% 
  inner_join(df.CUTI, by = "year") %>% 
  mutate(sar.method = "DART")

#NCBI
df.cbr.ncbi<-df.cbr %>% 
  inner_join(cope, by = c("year"= "Year")) %>% 
  mutate(sar.method = "DART")

#join all indices
df.cbr<-rbind(df.cbr.cui, df.cbr.cuti, df.cbr.ncbi) 


#add to sar_raw_updated with DART sar method with CUI,CUTI, NCBI
sar_raw_data_updated<-rbind(sar_raw_data_updated, df.cbr)
```

# CJS estimated survival




# save updated sar_raw_data

```{r}
#save updated sar_raw_data
usethis::use_data(sar_raw_data_updated, overwrite = TRUE)
```
 
