---
title: "SARforecastDLM_generate_forecast"
date: "Last update: `r format(Sys.Date())`"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    css: styles.css
---


   
Markdown includes:

- ATSAR tutorial for DLM with `MARSS` package.

- Code to prepare `SARforecastDLM_forecast` dataset. Once DLM is run, `usethis::use_data(SARforecastDLM_generate_forecast, overwrite = TRUE)` to add `plot_data` the data/ folder. Forecast of DLM with `MARSS` package. see `1964 to 2023` section for DLM code used. 

  - import function for [NOAA upwelling indice](https://oceanview.pfeg.noaa.gov/products/upwelling/intro) via [ERDAPP package](https://cran.r-project.org/web/packages/rerddap/vignettes/rerddap.html)

  - import function for SAR from Columbia Basin Research-- *pending url updated link for data import*
  


```{r load_libraries, warning=FALSE, message=FALSE}
library(tidyverse) #data wrangling
library(rerddap) #import data from ERDDAP
library(MARSS) #DLM
library(here) #file paths
library(car)
library(plotly)
```



# 1964 to 2005

## ATSAR tutorial
Searching online, it appears that most of the code needed is outlined in [DLM tutorial](https://atsa-es.github.io/atsa-labs/sec-dlm-fitting-a-univariate-dlm-with-marss.html) by Eli Holmes and Mark Scheurelle for ATSAR series. Has video tutorials as well. Uses the package `MARSS` developed by EH and MS with co-authors; require specific inputs for model to run.  [Ropensci/ATSA](https://atsa-es.r-universe.dev/packages). 



### `SalmonSurvCUI` data

Includes data from 1964 to 2005. See [Scheuerell and Williams (2005)](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1365-2419.2005.00346.x) for explanation of SAR for specific years. 
```{r import_data_MARSS}
## load the data
data(SalmonSurvCUI, package = "MARSS")

## get time indices
years <- SalmonSurvCUI[, 1]
## number of years of data
TT <- length(years)
## get response variable: logit(survival)
dat <- matrix(SalmonSurvCUI[,2], nrow = 1)
```

Using CUI with MARSS package, pull CUI and scale:
```{r z.score, eval=TRUE}
## get predictor variable: CUI
CUI <- SalmonSurvCUI[,3]
## z-score the CUI
CUI_z <- matrix((CUI - mean(CUI)) / sqrt(var(CUI)), nrow = 1)
## number of regr params (slope + intercept) = 2 
m <- dim(CUI_z)[1] + 1
```

```{r dlm-plotdata, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=4, fig.width=6}
par(mfrow = c(m, 1), mar = c(4, 4, 0.1, 0), oma = c(0, 0, 2, 0.5))
plot(years, dat, xlab = "", ylab = "Logit(SAR)",
     bty = "n", xaxt = "n", pch = 16, col = "darkgreen", type = "b")
plot(years, CUI_z, xlab = "", ylab = "CUI",
     bty = "n", xaxt = "n", pch = 16, col = "blue", type = "b")
axis(1, at = seq(1965, 2005, 5))
mtext("Year of ocean entry", 1, line = 3)
```

### build DLM

```{r univ.DLM.proc, eval=TRUE}
## for process eqn
B <- diag(m)                        ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)          ## 2x2; all 0 for now
diag(Q) <- c("q.alpha", "q.beta")   ## 2x2; diag = (q1,q2)
```

```{r univ.DLM.obs, eval=TRUE}
## for observation eqn
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1,1,] <- rep(1, TT)        ## Nx1; 1's for intercept
Z[1,2,] <- CUI_z             ## Nx1; predictor variable
A <- matrix(0)               ## 1x1; scalar = 0
R <- matrix("r")             ## 1x1; scalar = r
```


```{r univ.DLM.list, eval=TRUE}
## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
```

### fit DLM

```{r univ.DLM.fit, eval=TRUE}
## fit univariate DLM
dlm_1 <- MARSS(dat, inits = inits_list, model = mod_list)
```
### plot DLM

```{r dlm-plotdlm-1, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=4, fig.width=6}
ylabs <- c(expression(alpha[t]), expression(beta[t]))
colr <- c("darkgreen","blue")
par(mfrow = c(m, 1), mar = c(4, 4, 0.1, 0), oma = c(0, 0, 2, 0.5))
for(i in 1:m) {
  mn <- dlm_1$states[i,]
  se <- dlm_1$states.se[i,]
  plot(years, mn, xlab = "", ylab = ylabs[i], bty = "n", xaxt = "n", type = "n",
       ylim = c(min(mn-2*se), max(mn+2*se)))
  lines(years, rep(0,TT), lty = "dashed")
  lines(years, mn, col = colr[i], lwd = 3)
  lines(years, mn+2*se, col = colr[i])
  lines(years, mn-2*se, col = colr[i])
}
axis(1, at = seq(1965, 2005, 5))
mtext("Year of ocean entry", 1, line = 3)
```

### Forecasting

pulled from atsa chp 9.8:

Forecasting from a DLM involves two steps:

1. One-step ahead forecast (or prediction) of the regression parameters: Get an estimate of the regression parameters in time, t, from data up to time t-1.

2. One-step ahead forecast (or prediction) of the observation: Make a prediction of y at time t based on the predictor variables at time t and the estimate of the regression parameters at time t (step 1). 


```{r univ.DLM.fore_mean, eval=TRUE}
## get list of Kalman filter output
kf_out <- MARSSkfss(dlm_1)

## forecasts of regr parameters; 2xT matrix
eta <- kf_out$xtt1 #one step ahead forecast of regr parameters calc with kalman filter, represented as xtt1 in marss notation

## ts of E(forecasts)
fore_mean <- vector()
for(t in 1:TT) {
  fore_mean[t] <- Z[,,t] %*% eta[, t, drop = FALSE]
}
```


```{r univ.DLM.fore_var, eval=TRUE}
## variance of regr parameters; 1x2xT array
Phi <- kf_out$Vtt1 #the variances of the one-step ahead forecasts of the regression parameters at time t (Vtt1 in MARSS notation)

## obs variance; 1x1 matrix
R_est <- coef(dlm_1, type="matrix")$R

## ts of Var(forecasts)
fore_var <- vector()
for(t in 1:TT) {
  tZ <- matrix(Z[, , t], m, 1) ## transpose of Z
  fore_var[t] <- Z[, , t] %*% Phi[, , t] %*% tZ + R_est
}
```

### plot forecast


```{r dlm-plotdlmForeLogit, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=3, fig.width=6}
#logit scale
par(mar = c(4, 4, 0.1, 0), oma = c(0, 0, 2, 0.5))
ylims=c(min(fore_mean - 2*sqrt(fore_var)), max(fore_mean+2*sqrt(fore_var)))
plot(years, t(dat), type = "p", pch = 16, ylim = ylims,
     col = "blue", xlab = "", ylab = "Logit(SAR)", xaxt = "n")
lines(years, fore_mean, type = "l", xaxt = "n", ylab = "", lwd = 3)
lines(years, fore_mean+2*sqrt(fore_var))
lines(years, fore_mean-2*sqrt(fore_var))
axis(1, at = seq(1965, 2005, 5))
mtext("Year of ocean entry", 1, line = 3)
```

```{r dlm-plotdlmForeRaw, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=3, fig.width=6}
#inverse logit scale--back to SAR scale
invLogit <- function(x) {
  1 / ( 1 + exp(-x))
  }
ff <- invLogit(fore_mean)
fup <- invLogit(fore_mean+2*sqrt(fore_var))
flo <- invLogit(fore_mean-2*sqrt(fore_var))
par(mar = c(4, 4, 0.1, 0), oma = c(0, 0, 2, 0.5))
ylims <- c(min(flo), max(fup))
plot(years, invLogit(t(dat)), type = "p", pch = 16, ylim = ylims,
     col = "blue", xlab = "", ylab = "Survival (SAR)", xaxt = "n")
lines(years, ff, type = "l", xaxt = "n", ylab = "", lwd = 3)
lines(years, fup)
lines(years, flo)
axis(1, at = seq(1965, 2005, 5))
mtext("Year of ocean entry", 1, line = 3)
```

### diagnose forecast

```{r dlmInnov, eval=TRUE, echo=TRUE}
## forecast errors
innov <- kf_out$Innov
```

```{r dlmQQplot, eval=FALSE, echo=TRUE}
## Q-Q plot of innovations
qqnorm(t(innov), main = "", pch = 16, col = "blue")
## add y=x line for easier interpretation
qqline(t(innov))
```

```{r dlm-plotdlmQQ, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=2, fig.width=4}
## use layout to get nicer plots
layout(matrix(c(0,1,1,1,0),1,5,byrow=TRUE))

## set up L plotting space
par(mar=c(4,4,1,0), oma=c(0,0,0,0.5))

## Q-Q plot of innovations
qqnorm(t(innov), main="", pch=16, col="blue")
qqline(t(innov))
```

```{r dlmInnovTtest, eval=TRUE, echo=TRUE}
## p-value for t-test of H0: E(innov) = 0
t.test(t(innov), mu = 0)$p.value
```

```{r dlmACFplot, eval=FALSE, echo=TRUE}
## plot ACF of innovations
acf(t(innov), lag.max = 10)
```

```{r dlm-plotdlmACF, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=2, fig.width=4}
## use layout to get nicer plots
layout(matrix(c(0, 1, 1, 1, 0), 1, 5, byrow = TRUE))

## set up plotting space
par(mar = c(4, 4, 1, 0), oma = c(0, 0, 0, 0.5))

## ACF of innovations
acf(t(innov), lwd = 2, lag.max = 10)
```










# 1964 to 2023, CUI

## ERDAPP, `SalmonSurvCUI`, and CBR



### ERDAPP
NOAA upwelling indice CUI

```{r import_CUI_ERDAPP, message=FALSE}

# Set the URL of the ERDDAP server
url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdUI45mo", url = url)

# Download the dataset
dat <- griddap(info,
               time = c("1964-01-01T00:00:00Z", "2024-02-15T00:00:00Z"), #currently pulling all data to date (adjust as needed)
               latitude = c(45,45)
)

# format time and lat/long
df.CUI<-dat$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day), 
         adjlong = longitude-360) #set to -180 to 180

## only include april CUI from ERDAPP data (for now)
df.CUI<-df.CUI %>% 
  filter(month == 4) %>% 
  select(year, upwelling_index) 
  

```



### `SalmonSurvCUI`:CBR
Smolt-to-adult (SAR) survival via CBR appended to `SalmonSurvCUI` data.

Accessed wild Chinook Snake R Spring/Summer ESU SAR survival from Lower Granite (LGR) to Lower Granite (LGA), for in-river, transport, and bypass route fish. Excluded 0,1-Year Adult Detections per CBR recommendation. 

[CBR DART::PIT Tag Columbia Vasin ESU & DPS Queries](https://www.cbr.washington.edu/dart/query/pit_sar_esu)
```{r import_SAR_CBR, warning=FALSE, message=FALSE}
#CBR SAR start reporting in 2002; pull MARSS provided data from 1964 to 2002 
data(SalmonSurvCUI, package = "MARSS")

# keep year and logit.s; drop CUI
sar_64to99<-filter(SalmonSurvCUI, between(year, 1964,1999)) %>% 
  select(year, logit.s) 

#future: ask susannah about query html link to import directly into R for SAR csv. 
#For now, use previously generated .csv and import for SAR through present
W_SAR<-read.csv(here("data-raw/SAR","LGRtoLGA_allpass_W_spsu_ch.csv"))
W_SAR<-W_SAR[-c(24:31),] #drop CBR notes


#combine SAR data from 1964 to present with erdapp imported CUI

#dataset to test
sar.obs<-W_SAR %>%
  mutate(year = as.numeric(year)) %>%
  select(year, meanSAR) %>%
  mutate( logit.s = car::logit(meanSAR)) %>%
  select(year, logit.s) %>%
  rbind(sar_64to99) %>%
  left_join(df.CUI, by = c("year"="year")) %>%
  arrange(year) %>%
  filter(!is.na(logit.s))

#dataset to pred (includes CUI for 2022 and 2023 with no SAR). Confirm this method as "forecasting" and not just filling in NA values
sar.pred<-W_SAR %>% 
  mutate(year = as.numeric(year)) %>% 
  select(year, meanSAR) %>% 
  mutate( logit.s = logit(meanSAR)) %>% 
  select(year, logit.s) %>% 
  rbind(sar_64to99) %>% 
  left_join(df.CUI, by = c("year"="year")) %>% 
  arrange(year)  %>% 
  filter(!is.na(upwelling_index))



```


```{r extract_year_reponse_SAR}

#extract time indices, number of years of data, and response variable: logit(survival)

# extract year, tt,and response for 
years <- sar.obs[, 1]
## number of years of data
TT <- length(years)
## get response variable: logit(survival)
dat <- matrix(sar.obs[,2], nrow = 1)
  

```


```{r z.score-CUISAR, eval=TRUE}
## get predictor variable
CUI <- sar.obs[,3]
## z-score the CUI
CUI_z <- matrix((CUI - mean(CUI)) / sqrt(var(CUI)), nrow = 1)
## number of regr params (slope + intercept)
m <- dim(CUI_z)[1] + 1

```


```{r dlm-CUISAR-plotdata_plotly, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=4, fig.width=6}

plotly::ggplotly(
  ggplot(data = sar.pred, aes(x = year, y = round(logit.s,2))) +
  geom_point() +
  geom_line() +
  labs(title = "Logit(SAR) 1964 to 2021",
       x = "Year",
       y = "Logit(SAR)") +
    scale_x_continuous(breaks = seq(1964, 2024, 5)) +
  theme_minimal() 
)

plotly::ggplotly(
  ggplot(data = sar.pred, aes(x = year, y = CUI)) +
  geom_point() +
  geom_line() +
  labs(title = "CUI 1964 to 2023",
       x = "Year",
       y = "CUI") +
    scale_x_continuous(breaks = seq(1964, 2024, 5)) +
  theme_minimal() 
)

```



### build DLM 

```{r univ.dlm-CUISAR-.proc, eval=TRUE}
## for process eqn
B <- diag(m)                        ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)          ## 2x2; all 0 for now
diag(Q) <- c("q.alpha", "q.beta")   ## 2x2; diag = (q1,q2)
```

```{r univ.dlm-CUISAR-.obs, eval=TRUE}
## for observation eqn
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1,1,] <- rep(1, TT)        ## Nx1; 1's for intercept
Z[1,2,] <- CUI_z             ## Nx1; predictor variable
A <- matrix(0)               ## 1x1; scalar = 0
R <- matrix("r")             ## 1x1; scalar = r
```


```{r univ.dlm-CUISAR-.list, eval=TRUE}
## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
```

### fit DLM 
Includes NA for SAR 2022/2023 for forecasting.

```{r univ.dlm-CUISAR-.fit, eval=TRUE}
## fit univariate DLM
dlm_2 <- MARSS(dat, inits = inits_list, model = mod_list)
```
### plot DLM

```{r dlm-CUISAR--plotdlm-3, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=4, fig.width=6}
ylabs <- c(expression(alpha[t]), expression(beta[t]))
colr <- c("darkgreen","blue")
par(mfrow = c(m, 1), mar = c(4, 4, 0.1, 0), oma = c(0, 0, 2, 0.5))
for(i in 1:m) {
  mn <- dlm_2$states[i,]
  se <- dlm_2$states.se[i,]
  plot(years, mn, xlab = "", ylab = ylabs[i], bty = "n", xaxt = "n", type = "n",
       ylim = c(min(mn-2*se), max(mn+2*se)))
  lines(years, rep(0,TT), lty = "dashed")
  lines(years, mn, col = colr[i], lwd = 3)
  lines(years, mn+2*se, col = colr[i])
  lines(years, mn-2*se, col = colr[i])
}
axis(1, at = seq(1964, 2024, 5))
mtext("Year of ocean entry", 1, line = 3)
```

### forecasting DLM
https://atsa-es.github.io/MARSS/reference/predict_marssMLE.html
```{r}
fr<-MARSS::forecast(dlm_2, h= 2, type = "ytT", interval = "confidence", newdata = sar.pred)
plot(fr)

forecast_df$year <- (1964:2021)
forecast_df<-fr$pred
fr$newdata
df
fr$year

# add CUI for 2022/2023
new_values <- c(-17, -29)  # replace with actual values

# Create new Z matrix with only two time points
Z_new <- matrix(c(new_values), nrow = 1, ncol = 2, byrow = TRUE)

predict(dlm_2, newdata = sar.pred, type = "ytT")

matrix(5,2,10)
t
```

```{r univ.dlm-CUISAR-.fore_mean, eval=TRUE}
## get list of Kalman filter output
kf_out <- MARSSkfss(dlm_2)

## forecasts of regr parameters; 2xT matrix
eta <- kf_out$xtt1

## ts of E(forecasts)
fore_mean <- vector()
for(t in 1:TT) {
  fore_mean[t] <- Z[,,t] %*% eta[, t, drop = FALSE]
}
```


```{r univ.dlm-CUISAR-.fore_var, eval=TRUE}
## variance of regr parameters; 1x2xT array
Phi <- kf_out$Vtt1

## obs variance; 1x1 matrix
R_est <- coef(dlm_2, type="matrix")$R

## ts of Var(forecasts)
fore_var <- vector()
for(t in 1:TT) {
  tZ <- matrix(Z[, , t], m, 1) ## transpose of Z
  fore_var[t] <- Z[, , t] %*% Phi[, , t] %*% tZ + R_est
}
```


### plot(ly) forecast

```{r dlm-CUISAR-plotdata, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
#fct inv logit
invLogit <- function(x) {
  1 / ( 1 + exp(-x))
}


# Create a data frame for plotting from model --use for this markdown
plot_data_rmd <- data.frame(
  years = years, #years
  logit.s = t(dat), #observed sar data_scaled,
  sar = invLogit(t(dat)),  #observed sar data
  fore_mean_z = fore_mean, #mean of forecast scaled
  fore_var_z = fore_var, #variance of forecast scaled
  CUI_z = Z[1,2,] #predictor variable scaled 
)  %>% 
  mutate(
  ff = invLogit(fore_mean), # mean untransformedSAR mean forecast
  fupr = invLogit(fore_mean+2*sqrt(fore_var)), #upper 
  flwr = invLogit(fore_mean-2*sqrt(fore_var)) #lower
  )
```

```{r dlm-CUISAR--plotdlmForeLogit_plotly, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=3, fig.width=6}

# Create the plot (logit.s)
p<-ggplot(plot_data_rmd, aes(x = years)) +
  geom_point(aes(y = logit.s), color = "#b47747") +
  geom_line(aes(y = fore_mean_z)) +
  geom_line(aes(y = fore_mean_z + 2 * sqrt(fore_var_z)), linetype = "dashed") +
  geom_line(aes(y = fore_mean_z - 2 * sqrt(fore_var_z)), linetype = "dashed") +
  scale_x_continuous(breaks = seq(1964, 2024, 10)) +
  scale_linetype_manual(values = c("solid", "dashed"),
                        breaks = c("Forecasted logit.sar", "95% CI")) +
  labs(x = "Year of ocean entry", 
       y = "Logit(SAR)") + 
  # annotate("rect", xmin = 2021, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "grey", alpha = 0.2) +
  theme_classic() +
  theme(axis.line = element_line(color = "black"))
 
p_plotly<-ggplotly(p)

# Add rectangle
p_plotly <- p_plotly %>% 
  plotly::add_trace(
    x = c(2021, 2024, 2024, 2021), 
    y = c(-8.5, -8.5, -1.8, -1.8), 
    fill = "toself", 
    fillcolor = "grey", 
    line = list(width = 0), 
    opacity = 0.2, 
    showlegend = FALSE, 
    type = "scatter",
    mode = "none"
  )

p_plotly
```




```{r dlm-CUISAR--plotdlmForeRaw, eval=FALSE, echo=FALSE, fig=TRUE, fig.height=3, fig.width=6 }
#Code chunk was used for SAR plot featured on shinyAPP--pre full plotly transformation
p<-ggplot(plot_data_rmd, aes(x = years)) +
  geom_point(aes(y = sar), color = "#b47747") +
  geom_line(aes(y = ff)) +
  geom_line(aes(y = fupr), linetype = "dashed") +
  geom_line(aes(y = flwr), linetype = "dashed") +
  scale_x_continuous(breaks = seq(1964, 2024, 10)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "Year of ocean entry", 
       y = "Smolt-to-Adult Survival Ratio") + 
  # annotate("rect", xmin = 2021, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "grey", alpha = 0.2) +
  theme_classic() +
  theme(axis.line = element_line(color = "black"))


p_plotly<-ggplotly(p)

# Add rectangle
p_plotly <- p_plotly %>% 
  plotly::add_trace(
    x = c(2021, 2024, 2024, 2021), 
    y = c(0, 0, .13, .13), 
    fill = "toself", 
    fillcolor = "grey", 
    line = list(width = 0), 
    opacity = 0.2, 
    showlegend = FALSE, 
    type = "scatter",
    mode = "none"
  )

p_plotly
```



### diagnose forecast

```{r dlm-CUISAR-Innov, eval=TRUE, echo=TRUE}
## forecast errors
innov <- kf_out$Innov
```

```{r dlm-CUISAR-QQplot, eval=FALSE, echo=TRUE}
## Q-Q plot of innovations
qqnorm(t(innov), main = "", pch = 16, col = "blue")
## add y=x line for easier interpretation
qqline(t(innov))
```

```{r dlm-CUISAR--plotdlmQQ, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=2, fig.width=4}
## use layout to get nicer plots
layout(matrix(c(0,1,1,1,0),1,5,byrow=TRUE))

## set up L plotting space
par(mar=c(4,4,1,0), oma=c(0,0,0,0.5))

## Q-Q plot of innovations
qqnorm(t(innov), main="", pch=16, col="blue")
qqline(t(innov))
```
```{r dlm-CUISAR-InnovTtest, eval=TRUE, echo=TRUE}
## p-value for t-test of H0: E(innov) = 0
t.test(t(innov), mu = 0)$p.value
```

```{r dlm-CUISAR-ACFplot, eval=FALSE, echo=TRUE}
## plot ACF of innovations
acf(t(innov), lag.max = 10)
```

```{r dlm-CUISAR--plotdlmACF, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=2, fig.width=4}
## use layout to get nicer plots
layout(matrix(c(0, 1, 1, 1, 0), 1, 5, byrow = TRUE))

## set up plotting space
par(mar = c(4, 4, 1, 0), oma = c(0, 0, 0, 0.5))

## ACF of innovations
acf(t(innov), lwd = 2, lag.max = 10)
```






## append data
```{r append_plot_data_1, include=FALSE, eval=FALSE}
# save dataframe of only untransformed results for shinyapp
plot_data_1 <- data.frame(
  years = years, #years
  sar.obs = invLogit(t(dat)),  #observed sar data
  value = sar.pred[,3] #predictor variable CUI  
)  %>% 
  mutate(
  fore_mean_raw = invLogit(fore_mean), # mean untransformedSAR mean forecast
  fore_var_upper = invLogit(fore_mean+2*sqrt(fore_var)), #upper 
  fore_var_lower = invLogit(fore_mean-2*sqrt(fore_var)), #lower
  rear_type = "Both",
  index = "CUI",
  pass_type = "Both"
  )
```

# 1964 to 2023, CUTI

## ERDAPP, `SalmonSurvCUI`, and CBR



### ERDAPP
NOAA upwelling indice CUI



```{r import_CUTI_ERDAPP, message=FALSE}


# Get information about the erdCUTI dataset
info <- info(datasetid = "erdCUTImonthly", url = url)


# Download the dataset
dat <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2024-02-15T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)


df.CUTI<-dat$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day)) %>% 
  filter(month == 4)  #only include april 
  

```


### `SalmonSurvCUI`:CBR
Smolt-to-adult (SAR) survival via CBR appended to `SalmonSurvCUI` data.

Accessed wild Chinook Snake R Spring/Summer ESU SAR survival from Lower Granite (LGR) to Lower Granite (LGA), for in-river, transport, and bypass route fish. Excluded 0,1-Year Adult Detections per CBR recommendation. 

[CBR DART::PIT Tag Columbia Vasin ESU & DPS Queries](https://www.cbr.washington.edu/dart/query/pit_sar_esu)
```{r import_SAR_CBR_CUTI, warning=FALSE, message=FALSE}
#CBR SAR start reporting in 2002; pull MARSS provided data from 1988 to 1999
data(SalmonSurvCUI, package = "MARSS")

# CUTI only runs from 1988- keep year and logit.s; drop CUI
sar_88to99<-filter(SalmonSurvCUI, between(year, 1988,1999)) %>% 
  select(year, logit.s) 

#future: ask susannah about query html link to import directly into R for SAR csv. 
#For now, use previously generated .csv and import for SAR through present
W_SAR<-read.csv(here("data-raw/SAR","LGRtoLGA_allpass_W_spsu_ch.csv"))
W_SAR<-W_SAR[-c(24:31),] #drop CBR notes


#combine SAR data from 1964 to present with erdapp imported CUI

# #dataset to test
# sar.obs<-W_SAR %>% 
#   mutate(year = as.numeric(year)) %>% 
#   select(year, meanSAR) %>% 
#   mutate( logit.s = car::logit(meanSAR)) %>% 
#   select(year, logit.s) %>% 
#   rbind(sar_64to99) %>% 
#   left_join(df.CUI, by = c("year"="year")) %>% 
#   arrange(year) %>% 
#   filter(!is.na(logit.s))

#dataset to pred (includes CUTI for 2022 and 2023 with no SAR). Confirm this method as "forecasting" and not just filling in NA values
sar.pred<-W_SAR %>% 
  mutate(year = as.numeric(year)) %>% 
  select(year, meanSAR) %>% 
  mutate( logit.s = logit(meanSAR)) %>% 
  select(year, logit.s) %>% 
  rbind(sar_88to99) %>% 
  left_join(df.CUTI, by = c("year"="year")) %>% 
  arrange(year)  %>% 
  filter(!is.na(CUTI)) %>% 
  select(year, logit.s, CUTI)

```


```{r extract_year_reponse_SAR_CUTI}

#extract time indices, number of years of data, and response variable: logit(survival)

# extract year, tt,and response for 
years <- sar.pred[, 1]
## number of years of data
TT <- length(years)
## get response variable: logit(survival)
dat <- matrix(sar.pred[,2], nrow = 1)
  

```


```{r z.score-CUTISAR, eval=TRUE}
## get predictor variable
CUTI <- sar.pred[,3]
## z-score the CUTI
CUTI_z <- matrix((CUTI - mean(CUTI)) / sqrt(var(CUTI)), nrow = 1)
## number of regr params (slope + intercept)
m <- dim(CUTI_z)[1] + 1

```


```{r dlm-CUTISAR-plotdata_plotly, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=4, fig.width=6}

plotly::ggplotly(
  ggplot(data = sar.pred, aes(x = year, y = round(logit.s,2))) +
  geom_point() +
  geom_line() +
  labs(title = "Logit(SAR) 1988 to 2021",
       x = "Year",
       y = "Logit(SAR)") +
    scale_x_continuous(breaks = seq(1988, 2024, 5)) +
  theme_minimal() 
)


plotly::ggplotly(
  ggplot(data = sar.pred, aes(x = year, y = CUTI)) +
  geom_point() +
  geom_line() +
  labs(title = "CUTI 1988 to 2023",
       x = "Year",
       y = "CUTI") +
    scale_x_continuous(breaks = seq(1988, 2024, 5)) +
  theme_minimal() 
)

```



### build DLM 

```{r univ.dlm-CUTISAR-.proc, eval=TRUE}
## for process eqn
B <- diag(m)                        ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)          ## 2x2; all 0 for now
diag(Q) <- c("q.alpha", "q.beta")   ## 2x2; diag = (q1,q2)
```

```{r univ.dlm-CUTISAR-.obs, eval=TRUE}
## for observation eqn
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1,1,] <- rep(1, TT)        ## Nx1; 1's for intercept
Z[1,2,] <- CUTI_z             ## Nx1; predictor variable
A <- matrix(0)               ## 1x1; scalar = 0
R <- matrix("r")             ## 1x1; scalar = r
```


```{r univ.dlm-CUTISAR-.list, eval=TRUE}
## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
```

### fit DLM 
Includes NA for SAR 2022/2023 for forecasting.

```{r univ.dlm-CUTISAR-.fit, eval=TRUE}
## fit univariate DLM
dlm_3 <- MARSS(dat, inits = inits_list, model = mod_list)
```

### plot DLM

```{r dlm-CUTISAR--plotdlm-3, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=4, fig.width=6}
ylabs <- c(expression(alpha[t]), expression(beta[t]))
colr <- c("darkgreen","blue")
par(mfrow = c(m, 1), mar = c(4, 4, 0.1, 0), oma = c(0, 0, 2, 0.5))
for(i in 1:m) {
  mn <- dlm_3$states[i,]
  se <- dlm_3$states.se[i,]
  plot(years, mn, xlab = "", ylab = ylabs[i], bty = "n", xaxt = "n", type = "n",
       ylim = c(min(mn-2*se), max(mn+2*se)))
  lines(years, rep(0,TT), lty = "dashed")
  lines(years, mn, col = colr[i], lwd = 3)
  lines(years, mn+2*se, col = colr[i])
  lines(years, mn-2*se, col = colr[i])
}
axis(1, at = seq(1988, 2024, 5))
mtext("Year of ocean entry", 1, line = 3)
```

### forecasting DLM
```{r univ.dlm-CUTISAR-.fore_mean, eval=TRUE}
## get list of Kalman filter output
kf_out <- MARSSkfss(dlm_3)

## forecasts of regr parameters; 2xT matrix
eta <- kf_out$xtt1

## ts of E(forecasts)
fore_mean <- vector()
for(t in 1:TT) {
  fore_mean[t] <- Z[,,t] %*% eta[, t, drop = FALSE]
}
```


```{r univ.dlm-CUTISAR-.fore_var, eval=TRUE}
## variance of regr parameters; 1x2xT array
Phi <- kf_out$Vtt1

## obs variance; 1x1 matrix
R_est <- coef(dlm_3, type="matrix")$R

## ts of Var(forecasts)
fore_var <- vector()
for(t in 1:TT) {
  tZ <- matrix(Z[, , t], m, 1) ## transpose of Z
  fore_var[t] <- Z[, , t] %*% Phi[, , t] %*% tZ + R_est
}
```


### plot(ly) forecast

```{r dlm-CUTISAR-plotdata, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
# #fct inv logit
# invLogit <- function(x) {
#   1 / ( 1 + exp(-x))
# }


# Create a data frame for plotting from model --use for this markdown
plot_data_rmd <- data.frame(
  years = years, #years
  logit.s = t(dat), #observed sar data_scaled,
  sar = invLogit(t(dat)),  #observed sar data
  fore_mean_z = fore_mean, #mean of forecast scaled
  fore_var_z = fore_var, #variance of forecast scaled
  CUTI_z = Z[1,2,] #predictor variable scaled 
)  %>% 
  mutate(
  ff = invLogit(fore_mean), # mean untransformedSAR mean forecast
  fupr = invLogit(fore_mean+2*sqrt(fore_var)), #upper 
  flwr = invLogit(fore_mean-2*sqrt(fore_var)) #lower
  )
```

```{r dlm-CUTISAR--plotdlmForeLogit_plotly, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=3, fig.width=6}

# Create the plot (logit.s)
p<-ggplot(plot_data_rmd, aes(x = years)) +
  geom_point(aes(y = logit.s), color = "#b47747") +
  geom_line(aes(y = fore_mean_z)) +
  geom_line(aes(y = fore_mean_z + 2 * sqrt(fore_var_z)), linetype = "dashed") +
  geom_line(aes(y = fore_mean_z - 2 * sqrt(fore_var_z)), linetype = "dashed") +
  scale_x_continuous(breaks = seq(1988, 2024, 10)) +
  scale_linetype_manual(values = c("solid", "dashed"),
                        breaks = c("Forecasted logit.sar", "95% CI")) +
  labs(x = "Year of ocean entry", 
       y = "Logit(SAR)") + 
  # annotate("rect", xmin = 2021, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "grey", alpha = 0.2) +
  theme_classic() +
  theme(axis.line = element_line(color = "black"))
 
p_plotly<-ggplotly(p)

# Add rectangle
p_plotly <- p_plotly %>% 
  plotly::add_trace(
    x = c(2021, 2024, 2024, 2021), 
    y = c(-8.5, -8.5, -1.8, -1.8), 
    fill = "toself", 
    fillcolor = "grey", 
    line = list(width = 0), 
    opacity = 0.2, 
    showlegend = FALSE, 
    type = "scatter",
    mode = "none"
  )

p_plotly
```




```{r dlm-CUTISAR--plotdlmForeRaw, eval=FALSE, echo=FALSE, fig=TRUE, fig.height=3, fig.width=6 }
#Code chunk was used for SAR plot featured on shinyAPP--pre full plotly transformation
p<-ggplot(plot_data_rmd, aes(x = years)) +
  geom_point(aes(y = sar), color = "#b47747") +
  geom_line(aes(y = ff)) +
  geom_line(aes(y = fupr), linetype = "dashed") +
  geom_line(aes(y = flwr), linetype = "dashed") +
  scale_x_continuous(breaks = seq(1964, 2024, 10)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "Year of ocean entry", 
       y = "Smolt-to-Adult Survival Ratio") + 
  # annotate("rect", xmin = 2021, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "grey", alpha = 0.2) +
  theme_classic() +
  theme(axis.line = element_line(color = "black"))


p_plotly<-ggplotly(p)

# Add rectangle
p_plotly <- p_plotly %>% 
  plotly::add_trace(
    x = c(2021, 2024, 2024, 2021), 
    y = c(0, 0, .13, .13), 
    fill = "toself", 
    fillcolor = "grey", 
    line = list(width = 0), 
    opacity = 0.2, 
    showlegend = FALSE, 
    type = "scatter",
    mode = "none"
  )

p_plotly
```



### diagnose forecast

```{r dlm-CUISAR-Innov, eval=TRUE, echo=TRUE}
## forecast errors
innov <- kf_out$Innov
```

```{r dlm-CUISAR-QQplot, eval=FALSE, echo=TRUE}
## Q-Q plot of innovations
qqnorm(t(innov), main = "", pch = 16, col = "blue")
## add y=x line for easier interpretation
qqline(t(innov))
```

```{r dlm-CUISAR--plotdlmQQ, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=2, fig.width=4}
## use layout to get nicer plots
layout(matrix(c(0,1,1,1,0),1,5,byrow=TRUE))

## set up L plotting space
par(mar=c(4,4,1,0), oma=c(0,0,0,0.5))

## Q-Q plot of innovations
qqnorm(t(innov), main="", pch=16, col="blue")
qqline(t(innov))
```
```{r dlm-CUISAR-InnovTtest, eval=TRUE, echo=TRUE}
## p-value for t-test of H0: E(innov) = 0
t.test(t(innov), mu = 0)$p.value
```

```{r dlm-CUISAR-ACFplot, eval=FALSE, echo=TRUE}
## plot ACF of innovations
acf(t(innov), lag.max = 10)
```

```{r dlm-CUISAR--plotdlmACF, eval=TRUE, echo=FALSE, fig=TRUE, fig.height=2, fig.width=4}
## use layout to get nicer plots
layout(matrix(c(0, 1, 1, 1, 0), 1, 5, byrow = TRUE))

## set up plotting space
par(mar = c(4, 4, 1, 0), oma = c(0, 0, 0, 0.5))

## ACF of innovations
acf(t(innov), lwd = 2, lag.max = 10)
```






## append data
```{r append_plot_data_2, include=FALSE, eval=FALSE}
# save dataframe of only untransformed results for shinyapp
plot_data_2 <- data.frame(
  years = years, #years
  sar.obs = invLogit(t(dat)),  #observed sar data
  value = sar.pred[,3] #predictor variable CUI  
)  %>% 
  mutate(
  fore_mean_raw = invLogit(fore_mean), # mean untransformedSAR mean forecast
  fore_var_upper = invLogit(fore_mean+2*sqrt(fore_var)), #upper 
  fore_var_lower = invLogit(fore_mean-2*sqrt(fore_var)), #lower
  rear_type = "Both",
  index = "CUTI",
  pass_type = "Both"
  )
```



# save plot_data

```{r}

plot_data<-rbind(plot_data_1, plot_data_2)

#add to data/ folder 
usethis::use_data(plot_data, overwrite = TRUE)

```

