---
title: "compare model fit"
output: html_document
date: "2024-04-22"
---
```{r load_libraries, warning=FALSE, message=FALSE}
library(tidyverse) #data wrangling
library(rerddap) #import data from ERDDAP
library(MARSS) #DLM
library(here) #file paths
library(car)
library(plotly)
```


Objective: allow user to select a year of interest that filters the original data from 1964 to that year. Then uses those designated years as a training data set. it will then use the remainder of the years through 2005 as a the new test data and then plot 

ex. user selects 1964 to 1980, those years are now the train data and the rest are used as new data for testing

_base_plot_data: only includes S&W05 method now with wild, and all passage types? 
SalmonSurv: append CUI and CUTI

```{r}
base_plot_data<-base_plot_data %>% 
  rename("value" = "upwelling_index")

usethis::use_data(base_plot_data, overwrite = TRUE)

```




### `SalmonSurvCUI` data

```{r}

## load the data
data(SalmonSurvCUI, package = "MARSS")

#load CUTI
# Set the URL of the ERDDAP server
url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdCUTImonthly", url = url)


# Download the dataset
dat <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2024-02-15T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)


df.CUTI<-dat$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day)) %>% 
  filter(month == 4) %>% 
  select(year, CUTI)  %>% 
  rename(value = "CUTI") %>% 
  mutate(index = "CUTI")

df<-SalmonSurvCUI %>% 
  select(year, logit.s)

df.cuti<-df %>% inner_join( df.CUTI, by = "year")
  
#join data - returns year and sar repeated for cui and cuti
sar_raw_data<- SalmonSurvCUI %>% 
 rename("value" = CUI.apr) %>% 
  mutate(index = "CUI") %>% 
  bind_rows(df.cuti)

usethis::use_data(sar_raw_data)
  
```


#### CUI
Includes data from 1964 to 2005. See [Scheuerell and Williams (2005)](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1365-2419.2005.00346.x) for explanation of SAR for specific years. 
```{r}
library(tidyverse)

data<- sar_raw_data

year_selected<-1995

index_selected<-"CUI"
```


```{r function}
fct_compare_model_years<-function(data, years_selected, index_selected){

full_data<- data %>% 
  dplyr::filter(between(year, min(year), 2005),
                index == index_selected)

train_data<- full_data %>%
  dplyr::filter(between(year, min(year),max(years_selected)), 
                index == index_selected) 

# test_data<- full_data %>%
# dplyr::filter(between(year,2000 +1, 2005),#max(years_selected)
#               index == "CUI")#index_selected) 
  
## get time indices
years <- train_data$year
## number of years of data
TT <- length(years)
## get response variable: logit(survival)
dat <- matrix(train_data$logit.s, nrow = 1)

## get predictor variable
index <- full_data$value

## ## ## z-score the upwelling index
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)

index_z_train <- index_z[1:TT]

## number of regr params (slope + intercept) = 2 
m <- dim(index_z)[1] + 1


### build DLM

## for process eqn
B <- diag(m)                        ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)          ## 2x2; all 0 for now
diag(Q) <- c("q.alpha", "q.beta")   ## 2x2; diag = (q1,q2)

## for observation eqn
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1,1,] <- rep(1, TT)        ## Nx1; 1's for intercept
Z[1,2,] <- index_z_train             ## Nx1; predictor variable
A <- matrix(0)               ## 1x1; scalar = 0
R <- matrix("r")             ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)


### fit DLM
set.seed(1234)
## fit univariate DLM
dlm_train <- MARSS(dat, inits = inits_list, model = mod_list)


#pull indice with full data z score
index_z_test <- index_z[(TT+1)] #get 3 CUI ahead

#newdata z array
Z_test <- array(NA, c(1, m, 3))  ## NxMxT; empty for now
Z_test[1,1,] <- rep(1, 3)        ## Nx1; 1's for intercept
Z_test[1,2,] <- index_z_test

## forecast
forecast_df<-forecast(dlm_train, h = 1, type = "ytT", interval = "confidence")

#extract forecast
forecast_df_pred<- forecast_df$pred


#wrangle for plot
forecast_df_raw<-forecast_df_pred%>% 
  mutate(across(c(,3:9), ~plogis(.) * 100)) %>% 
  rename("fore_CI_95_upper" = `Hi 95`,
         "fore_CI_95_lower" = `Lo 95`) %>% 
  mutate(year = (min(years):max(years+1)),
         index = index_selected,
         sar.method = "Scheuerell and Williams (2005)",
         dataset = "select_forecast",
         rear_type = "Natural-origin",
         pass_type = "All") %>% 
  inner_join(select(full_data, value, year), by = "year") %>% ##join index value
  select(year, value, y, estimate, se, fore_CI_95_lower, fore_CI_95_upper,sar.method,index, rear_type, pass_type, dataset)


return(forecast_df_raw)
}
```


```{r run_function}
df_raw<-fct_compare_model_years(data = data, years_selected = year_selected, index_selected = index_selected)

df_raw1<-fct_model_forecast(data = data, years_selected = year_selected, index_selected = index_selected)

fct_forecast_compare_plot(data_base = base_plot_data, data_select = df_raw1, years_selected = year_selected)

```

```{r fct_adding new data}
library(tidyverse)
library(MARSS)
#ex if user selected 1990:

#split data test.train.full
full_data<- SalmonSurvCUI %>% 
 rename("value" = CUI.apr) %>% 
  mutate(index = "CUI")
  

fct_testtrain_year<-function(data, years_selected = 1990, index_selected = "CUI"){

full_data<- data

train_data<- full_data %>%
  dplyr::filter(between(year, 1964, max(years_selected)), 
                index == index_selected) 

test_data<- full_data %>%
dplyr::filter(between(year, max(years_selected)+1, 2005),
              index == index_selected) 
  
## get time indices
years <- train_data[, 1]
## number of years of data
TT <- length(years)
## get response variable: logit(survival)
dat <- matrix(train_data[,2], nrow = 1)

## get predictor variable: CUI
index <- full_data[,3]

## ## ## z-score the CUI
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)

index_z_train <- index_z[1:TT]

## number of regr params (slope + intercept) = 2 
m <- dim(index_z)[1] + 1


### build DLM

## for process eqn
B <- diag(m)                        ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)          ## 2x2; all 0 for now
diag(Q) <- c("q.alpha", "q.beta")   ## 2x2; diag = (q1,q2)

## for observation eqn
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1,1,] <- rep(1, TT)        ## Nx1; 1's for intercept
Z[1,2,] <- index_z_train             ## Nx1; predictor variable
A <- matrix(0)               ## 1x1; scalar = 0
R <- matrix("r")             ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)


### fit DLM

## fit univariate DLM
dlm_train <- MARSS(dat, inits = inits_list, model = mod_list)


#reformat test data to use as newdata
#test response data
y_new<-matrix(test_data[,2], nrow = 1)

#length of forecast for remaining years
testTT<- length(y_new)

#pull indice with full data z score
index_z_test <- index_z[(TT+1):42]

#newdata z array
Z_test <- array(NA, c(1, m, testTT ))  ## NxMxT; empty for now
Z_test[1,1,] <- rep(1, testTT)        ## Nx1; 1's for intercept
Z_test[1,2,] <- index_z_test

## forecast
forecast_df<-forecast(dlm_train, h = 15, type = "ytT", interval = "confidence", newdata = list(y = y_new , Z = Z_test))

#extract forecast
forecast_df_pred<- forecast_df$pred


#wrangle for plot
forecast_df_raw<-forecast_df_pred%>% 
  mutate(across(c(,3:9), ~plogis(.) * 100)) %>% 
  rename("fore_CI_95_upper" = `Hi 95`,
         "fore_CI_95_lower" = `Lo 95`) %>% 
  mutate(year = 1964:2005,
         index = "CUI", 
         sar.method = "Scheuerell and Williams (2005)",
         dataset = if_else(year %in% train_data$year, "Train", "Test"),
         rear_type = "Natural-origin",
         pass_type = "All") %>% 
  inner_join(select(full_data, value, year), by = "year") 


return(forecast_df_raw)
}

df_raw<-fct_testtrain_year(data = full_data)

fct_forecast_plot(data=df_raw) 

```


```{r dlm1_model}
## load the data
data(SalmonSurvCUI, package = "MARSS")

## get time indices
years <- SalmonSurvCUI[, 1]
## number of years of data
TT <- length(years)
## get response variable: logit(survival)
dat <- matrix(SalmonSurvCUI[,2], nrow = 1)

## get predictor variable: CUI
CUI <- SalmonSurvCUI[,3]

## ## ## z-score the CUI
CUI_z <- matrix((CUI - mean(CUI)) / sqrt(var(CUI)), nrow = 1)

## number of regr params (slope + intercept) = 2 
m <- dim(CUI_z)[1] + 1

### build DLM

## for process eqn
B <- diag(m)                        ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)          ## 2x2; all 0 for now
diag(Q) <- c("q.alpha", "q.beta")   ## 2x2; diag = (q1,q2)

## for observation eqn
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1,1,] <- rep(1, TT)        ## Nx1; 1's for intercept
Z[1,2,] <- CUI_z             ## Nx1; predictor variable
A <- matrix(0)               ## 1x1; scalar = 0
R <- matrix("r")             ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)


### fit DLM

## fit univariate DLM
dlm_1 <- MARSS(dat, inits = inits_list, model = mod_list)


forecast_df <- forecast(dlm_1, h = 2, type = "ytT", interval = "confidence")

plot(forecast_df)
```

```{r dlm1_forecast}
# Forecast the additional two years
forecast_df <- forecast(dlm_1, h = 2, type = "ytT", interval = "confidence")
# Print the forecast results
print(forecast_df)

plot(forecast_df)

forecast_df<- forecast_df$pred
forecast_df$year <- 1964:2007
forecast_df$index <- "Scheuerell and Williams (2005)"

accuracy(dlm_1)

# Define the logistic function
logistic <- function(x) {
  1 / (1 + exp(-x))
}

forecast_df_raw<-forecast_df%>% 
  mutate(across(c(,3:9), ~plogis(.) * 100)) %>% 
  rename("fore_CI_95_upper" = `Hi 95`,
         "fore_CI_95_lower" = `Lo 95`) %>% 
  mutate(year = 1964:2007,
         sar.method = "Scheuerell and Williams (2005)",
         index = "CUI", 
         rear_type = "Natural-origin",
         pass_type = "All") %>% 
  inner_join(df.CUI, by = "year") 

# #Check sd
# # Print the mean and standard deviation of the original standardized CUI data
# print(mean(CUI_z))
# print(sd(CUI_z))
# 
# # Print the mean and standard deviation of the new standardized CUI data
# print(mean(new_CUI_z))
# print(sd(new_CUI_z))
# 
# # Combine the original and new standardized CUI data
# combined_CUI_z <- c(CUI_z, new_CUI_z)
# 
# # Print the mean and standard deviation of the combined standardized CUI data
# print(mean(combined_CUI_z))
# print(sd(combined_CUI_z))

```


#### CUTI
```{r CUTI}
# Get information about the erdCUTI dataset
info <- info(datasetid = "erdCUTImonthly", url = url)


# Download the dataset
dat <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2024-02-15T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)


df.CUTI<-dat$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day)) %>% 
  filter(month == 4) %>% 
  select(year, CUTI)  

  
```

```{r dlm1CUTI_model}
## load the data
data(SalmonSurvCUI, package = "MARSS")

SalmonSurvCUTI<-SalmonSurvCUI %>% 
  filter(between(year, 1988, 2005))

## get time indices
years <- SalmonSurvCUTI[, 1]
## number of years of data
TT <- length(years)
## get response variable: logit(survival)
dat <- matrix(SalmonSurvCUTI[,2], nrow = 1)

## get predictor variable: CUTI
CUTI<-df.CUTI %>% filter(between(year, min(year),2005))

CUTI <- CUTI[,3]

## ## ## z-score the CUI
CUTI_z <- matrix((CUTI - mean(CUTI)) / sqrt(var(CUTI)), nrow = 1)

## number of regr params (slope + intercept) = 2 
m <- dim(CUTI_z)[1] + 1

### build DLM

## for process eqn
B <- diag(m)                        ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)          ## 2x2; all 0 for now
diag(Q) <- c("q.alpha", "q.beta")   ## 2x2; diag = (q1,q2)

## for observation eqn
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1,1,] <- rep(1, TT)        ## Nx1; 1's for intercept
Z[1,2,] <- CUTI_z             ## Nx1; predictor variable
A <- matrix(0)               ## 1x1; scalar = 0
R <- matrix("r")             ## 1x1; scalar = r

## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))

## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)


### fit DLM

## fit univariate DLM
dlm_1_CUTI <- MARSS(dat, inits = inits_list, model = mod_list)
```

```{r dlm1CUTI_forecast}
# Forecast the additional two years
forecastCUTI_df <- forecast(dlm_1_CUTI, h = 2, type = "ytT", interval = "confidence")
# Print the forecast results
print(forecastCUTI_df)

plot(forecastCUTI_df)


accuracy(dlm_1_CUTI)

forecastCUTI_df_raw<-forecastCUTI_df%>% 
  mutate(across(c(,3:9), ~stats::plogis(.) * 100)) %>% 
  rename("fore_CI_95_upper" = `Hi 95`,
         "fore_CI_95_lower" = `Lo 95`) %>% 
  mutate(year = 1988:2007,
         sar.method = "Scheuerell and Williams (2005)",
         index = "CUTI", 
         rear_type = "Natural-origin",
         pass_type = "All") %>% 
  inner_join(df.CUTI, by = "year") %>% 
  rename("upwelling_index" = CUTI)

```




